Learning how to act: making good decisions with machine learning

### 2017-10-22 thesis/thesis_map.tex
fixes typos

There are many people without whose support this thesis would not have been possible. I would like to thank my supervisors Dr Cheng Song Ong, Dr Mark Reid and Dr Tiberio Caetano, Prof. Robert Williamson for chairing my panel, my brother Tor for many interesting and insightful discussions on bandit problems, and my parents for helping me proofread and the many days given to care for my lovely daughters. I would also like to thank Victoria, Inger, Natalie and all the people at ANU thesis bootcamp for giving me the techniques I need to get past the blank page and get words out. Finally I want to thank my wonderful husband for his unwavering support and willingness to take on the household chaos, and my two beautiful daughters Freya and Anouk for their patience while mummy worked on her "tesis".

@@

\section{Motivation}
Many of the most important questions in science, commerce and our personal lives are about the outcomes of doing something. Will asking people to pay upfront at the doctors reduce long term health expenditure? If we developed a drug to suppress particular genes, could we cure multiple sclerosis? Would delaying teen-aged pregnancies improve the outcome for their children.

@@

The notion of causality has been widely debated in science and philosophy \citep{hume1741treatise,mill1893system,pearson1911grammar,Pearl2000,Lewis2000,woodward2005making,heckman20051,cartwright2007hunting} but is still widely viewed as poorly defined. This has led to a reluctance among applied researchers in many fields to make concrete claims about causality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}. However, the magnitude, direction and even existence of an association depends on which variables are controlled for (or included in a regression). Avoiding formalising causation, which is the real question of interest, requires the reader to determine via \quotes{common sense} the implications of the reported associations.

There are two ways in which an association detected in a data set may be non-causal. The first is that the variables concerned may not be related at all, and the association has arisen by chance in that data sample. Given a finite data on enough variables, there is a high probability of finding some that appear correlated even though they are completely unrelated. For example, based on data from the years 1999 to 2009, the age of Miss America is strongly correlated with the number of murders  (in the US) by steam, hot vapours and hot objects \citep{Vigen2015}. We would not expect this relationship to hold in a new sample of data. This form of spurious correlation also has serious repercussions. It lies at the heart of major problems with the process of scientific research: researchers are incentivised to detect effects and thus to explore many possible paths in the process of analysing data and studies that fail to find an effect are less likely to be published. Consequently, the likelihood that reported effects have arisen by chance is underestimated, leading to the conclusion that \quotes{most published scientific results are false} \citep{Ioannidis2005}. This issue is also highlighted by recent crises in replication \citep{OpenScienceCollaboration2015}. This issue can be ameliorated by obtaining more data and by separating learning models from evaluating their performance, for example by evaluating models on a strict hold-out set or on the extent to which their results can be replicated.

@@

In chapter \ref{chap:causal-bandit}, I introduce causal bandit problems that unify causal graphical models and multi-armed bandit problems. Bandit arms are related to interventions in a causal graphical model in a very natural way: each multi-armed bandit arm (or action) corresponds to a particular assignment of values to variables within the causal graphical model. I show how this framework can be used to describe a number of existing problems that lie in the intersection between the observational and interventional approaches to causality and demonstrate how problems reduce to different flavours of the bandit problem depending what information is observable and when.

In \S\ref{sec:causal-bandit-post-action-feedback}, I focus on causal bandit problems for which the values of variables in the causal graph are observed after an action is selected. I demonstrate that this leads to a novel form of structure between the bandit arms that cannot be exploited by any existing bandit algorithms. In \S\ref{sec:simple-regret}, I describe and develop an algorithm for a special case of the causal bandit problem, which I refer to as the \emph{parallel bandit problem}. I demonstrate via upper and lower bounds on the regret that the algorithm is close to optimal for these problems and that the introduction of the causal structure leads to substantially faster learning. In \S\ref{sec:simple-regret-general}, I develop and prove regret bounds for an algorithm that can be applied to arbitrary causal graphs, albeit with stronger assumptions on what must be known a priori, and introduce a measure that captures the underlying difficulty of causal bandit problems, which depends on the causal graph and can be viewed as an \quotes{effective number of arms}. I also show how this measure can be used to quantify the value of optimised interventional data over purely observational data. Section \ref{sec:causal-bandit-experiments} provides experiments demonstrating the performance of both the parallel and general causal bandit algorithms on causal bandit problems. Finally in \S\ref{sec:causal_bandit_discussion}, I discuss extensions and potential future work.

@@

Counterfactuals \citep{Rubin1974} were developed from the starting point of generalising from randomised trials to less controlled settings. They describe causal effects in terms of differences between counterfactual variables, what would happen if we took one action versus what would happen if we took another. Counterfactual assertions can be expressed very naturally in human languages and are  prevalent in  everyday conversations; \quotes{if I had worked harder I would have got better grades} and \quotes{she would have been much sicker if she hadn't taken antibiotics}. Structural equation models have been developed and applied primarily within economics and related disciplines. They can be seen as an attempt to capture key aspects of the people's behaviour with mathematics. Questions around designing policies or interventions play a central role in economics. Thus they have transformed simultaneous equations into a powerful (though somewhat poorly formalised \citep{Pearl2000}) framework and associated set of methods for estimating causal effects. Causal Bayesian networks \citep{Pearl2000} are a more recent development and arise from the addition of a fundamental assumption about the meaning of a link to Bayesian networks. They inherit and leverage the way Bayesian networks encode conditional independencies between variables to localise the impact of an intervention in a system in a way that allows formalisation of the conditions under which causal effects can be inferred from observational data.

An understanding of causal Bayesian networks and their properties (in particular the do calculus, see section \ref{sec:do_calculus}) is sufficient to appreciate my main technical contributions in chapter \ref{chap:causal-bandit}, as well as the importance of formal causal reasoning in Bayesian inference that I highlight in section  \ref{sec:bayesian_causality}. However, the literature on causal inference techniques remains split between the different frameworks. Much of the recent work on estimating causal effects within machine learning, as well as widely used methodologies such as propensity scoring, are described using the counterfactual framework. Methods developed within economics, in particular instrumental variable based approaches, or those requiring parametric or functional assumptions, are often based around structural equation models. This makes it worthwhile for researchers interested in causality to develop an understanding off all these viewpoints.

@@

\vspace*{.3cm}
\begin{example}
\label{exm:adjusting}
Suppose we want to estimate the impact on high school graduation rates of compulsory preschool for all four year olds. We have a large cross-sectional data set on a group of twenty year olds that records if they attended pre-school, if they graduated high school and their parents socio-economic status (SES). We will let $X\in \set{0,1}$ indicate if an individual attended pre-school, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.\footnote{There has been substantial empirical work on the effectiveness of early childhood education including a landmark randomised trial, the Perry Preschool project, which ran from 1962-1967 \citep{weikart1970longitudinal}.}
\end{example}

@@

Returning to example \ref{exm:ranomized_experiment}, and phrasing our query in terms of interventions; what would the distribution of outcomes look like if everyone was treated $\P{Y|do(X=1)}$, relative to if no one was treated $\P{Y|do(X=0)}$? The treatment $X$ is a potential cause of $Y$, along with other unobserved variables, such as the age, gender and the disease sub-type of the patient. Since $X$ is assigned via deliberate randomisation, it cannot be affected by any latent variables. The causal Bayesian network for this scenario is shown in figure \ref{fig:causal_network_example}. This network represents the (causal) factorisation  $\P{X,Y} = \P{X}\P{Y|X}$, so from equation (\ref{eq:truncatedproduct}), $\P{Y|do(X)} = \P{Y|X}$. In this example, the interventional distribution is equivalent to the observational one.

@@

Putting aside any estimates of individual causal effects, it is possible to learn something about the distributions under treatment or placebo. Let $\cf{Y}{1}$ be a random variable representing the potential outcome if treated. The distribution of $\cf{Y}{1}$ is the distribution of $Y$ if everyone was treated. Similarly $Y^{0}$ represents the potential outcome for the placebo. The difference between the probability of recovery, across the population, if everyone was treated and the probability of recovery given placebo is $\P{\cf{Y}{1}}-\P{\cf{Y}{0}}$. We can estimate (from an experimental or observational study):
\begin{itemize}
\item $\P{Y=1|X=1}$, the probability that those who took the treatment will recover
\item $\P{Y=1|X=0}$, the probability that those who were \emph{not} treated will recover
\end{itemize}

@@

The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignorability assumptions \citep{Rosenbaum1983}. They state that the treatment each person receives is independent of whether they would recover if treated and if they would recover if not treated. This is justified in example \ref{exm:ranomized_experiment} due to the randomisation of treatment assignment. In general the treatment assignment will not be independent of the potential outcomes. In example \ref{exm:adjusting}, the children who attended preschool may be more likely to have graduated high school had they in fact not attended than the children who actually did not attend, i.e. $X \nci \cf{Y}{0}$. Similarly, had the poorer children who did not attend pre-school attended they might not have done as well as the children who did in fact attend, i.e. $X \nci \cf{Y}{1}$. A more general form of the ignorability assumption is to identify a set of variables $Z$ such that $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$.

@@

Assuming that within each socio-economic status level, attendance at pre-school is independent of the likelihood of graduating high-school had a person attended, then the average rate of high-school graduation given a universal pre-school program can be computed from equation \ref{eqn:counterfactual1}. Note, that this agrees with the weighted adjustment formula in equation \ref{eqn:backdoor_example}.

Another assumption introduced within the Neyman-Rubin causal framework is the Stable Unit Treatment Value Assumption (SUTVA) \citep{Rubin1978}. This is the assumption that the potential outcome for one individual (or unit) does not depend on the treatment assigned to another individual. As an example of a SUTVA violation, suppose disadvantaged four year olds were randomly assigned to attend pre-school. The subsequent school results of children in the control group, who did not attend, could be boosted by the improved behaviour of those who did and who now share the classroom with them. SUTVA violations would manifest as a form of model misspecification in causal Bayesian networks.

@@

The SEM for example \ref{exm:adjusting} is shown in figure \ref{fig:sem:preschool}. Intervening to send all children to pre-school replaces the equation $X = f_x(Z,\epsilon_x)$ with $X=1$, leaving all the other functions and distributions in the model unchanged.

@@

\vspace{.3cm}
\begin{theorem}
\textbf{(Local Markov condition)} Given a Bayesian network $G$ with nodes $X_{1}...X_{N}$, each variable $X_{i}$ is independent of its non-decedents given its parents in $G$ for all distributions $P(X_{1}...X_{N})$ that are compatible with $G$.
\end{theorem}

@@

In some cases, the average causal effect for some sub-group of the population is of prime interest. A particularly common example of this is the average treatment effect of the treatment of the treated (ATT). This would be the key quantity of interest if we had to decide whether or not to continue providing a program or treatment for which we could not control the treatment assignment process.

@@

Assuming $x$ and $\vb{z}$ are discrete, equation \ref{eqn:expected_post_intervention}, and thus the ACE, can be estimated by selecting the data for which $X=x$, stratifying by $\vb{Z}$, then computing the mean outcome within each stratum and finally weighting the results by the number of samples in each strata. However this approach is not workable for most real world problems with finite samples as the number of strata grows exponentially with the dimension of $\vb{Z}$ and it cannot handle continuous covariates. There is a substantial body of work within in the statistics and econometrics literature on estimating average causal effects assuming conditional ignorability (see \citet{Imbens2004} for a comprehensive review). The key approaches are based on matching on covariates, propensity score methods and regression. We now examine these approaches from a machine learning perceptive.

@@

\subsubsection{Propensity scores}
If the parametric model is miss-specified then the parameter that minimises the loss depends on the distribution from which the covariates $z$ are sampled. The model learned by ERM could perform very well in a validation set (which estimates the generalisation error over the observational distribution of $(x,z)$) but yield very poor estimates of the causal effect, see figure \ref{fig:covariate_shift_adjust}.

\begin{figure}[ht]
\centering
\includegraphics[scale=.48]{covariate_shift_estimation.pdf}
\caption{
Parametric regression may yield poor estimates of causal effects if the model is miss-specified, even if the model fits well over the domain of the training data. In this example, $\P{Z|X=0} \sim N(\mu_0,\sigma_0^2)$ and $\P{Z|X=1} \sim N(\mu_1,\sigma_1^2)$ with little overlap in the densities. If $X=0$ then $Y \sim N(f_1(x)=sin(x),\sigma_y^2)$ and if $X=1$ then $Y \sim N(f_0(x) = \frac{1}{x+1},\sigma_y^2)$. We estimate $f_1(z)$ from the sample in which $X=1$ (green points) and $f_0(z)$ from the sample for which $X=0$ (blue points). In both cases the linear model is a good fit to the data. However, the resulting estimate of the causal effect is very poor for the lower values of $z$. } 
\label{fig:covariate_shift_adjust}
\end{figure}

@@

The estimator in equation \ref{eqn:importance_weighted_erm} is an example of a doubly robust estimator \c%fitep{Robins1994a,Kang2007}. Doubly robust methods are asymptotically unbiased so long as either the regression model $h$ or propensity score $e$ are correctly specified \citep{robins1995semiparametric}.

@@

This formulation of the regression estimator highlights the missing data aspect of casual effect estimation. For each instance, the regression models are used to estimate the counterfactual outcome had the instance received the alternate treatment. Matching estimates the counterfactual outcome for an instance from the outcome of \emph{similar} instances that received a different treatment. \citet{abadie2002simple} analyse an estimator where both target and control instances are matched and the matching is done with replacement, let $j \in J_k(i)$ be the indices of the $k$ instances closest to $i$ by some metric $d(z_i,z_j)$ such that $x_i \neq x_j$.

@@

The significance of this should not be underestimated. Cross-validation has allowed applied machine learning to succeed with a very a theoretical approach on the basis that we can identify when a model is successful. With causal effect estimation there is no guarantee that a model that performs well at prediction (even out of sample) will accurately estimate the outcome of an intervention. \citet{Sugiyama2007} propose inverse propensity weighted cross validation for the covariate shift problem. There is relatively little theory on model selection for estimating the propensity score. To achieve asymptotically unbiased estimates, the covariates should satisfy the backdoor criterion. It is also known that conditioning on instrumental variables, which directly influence $X$ but not $Y$, increases variance without any reduction in bias and can increase bias if there are unmeasured confounding variables \citep{Wooldridge2009,Bhattacharya2012,Pearl2012b,Myers2011}. With doubly robust estimators, one could apply an iterative approach, fitting a propensity score model, using the results for inverse propensity weighted cross-validation of the regression model and then selecting covariates for the propensity model on the assumption the estimated regression function was correct. I have found no examples of such an approach.

@@

Suppose data is generated by the linear Gaussian Bayesian network below. The goal is to estimate the causal effect of $X$ on $Y$, that is identifying the value of the coefficient $w_{yx}$. The variable $U$ is latent.

@@

where

\eqn{
\beta &= w_{yz}+\frac{w_{yu}w_{zu}}{w_{zu}^2+\frac{v_z}{v_u}}, \text{ and}\\
\epsilon &= v_y + w_{yu}^2v_u - \frac{v_u^2 w_{zu}^2 w_{yu}^2 }{v_z^2+v_u^2 w_{zu}^2}
}

@@

The previous sections focused on aspects of the problem of estimating the likely effect of an intervention from data gathered prior to making the intervention. There is an obvious alternative. Instead of trying to infer the outcome of an intervention from passive observations, one can intervene and see what happens. There are three key differences between observing a system and explicitly intervening in it. First, we determine the nature of the intervention and thereby control the data points used to estimate causal effects. Selecting data points optimally for learning is the focus of the optimal experimental design literature within statistics \citep{pukelsheim2006optimal} and the active learning literature in machine learning \citep{settles2010active}. Secondly, explicitly choosing interventions yields a perfect model of the probability with which each action is selected, given any context, allowing control over confounding bias. Finally, when we are intervening in a system we typically care about the impact of our actions on the system in addition to optimising learning. For example, in a drug trial , assigning people a sub-optimal treatment has real world costs. This leads to a trade-off between exploiting the best known action so far and exploring alternative actions about which we are less certain. This exploration-exploitation trade-off lies at the heart of the field of reinforcement learning \citep{sutton1998reinforcement}.

@@

We focus on a simpler class of problems within reinforcement learning known as multi-armed bandit problems. Multi-armed bandit problems also describe an agent aiming to maximise some reward by interacting with an environment. However, the reward is observed immediately after the action is selected and the environment is stateless. Future rewards generated by the  environment in response to the agents actions do not depend on the previous actions of the agent. These assumptions are a reasonable approximation in settings where we face the same decision repeatedly with respect to many independent units or individuals and we have a good proxy for the outcome. In this chapter we describe the interventionalist approaches to learning to act for these types of problems, beginning with classical randomised experiments and then the, more adaptive, multi-armed bandit algorithms. We show how both observational causal inference and bandit algorithms can be viewed as generalisations of randomised experiments and highlight connections between the observational and interventional approaches to causal inference.

@@

Stratified randomised experiments address the issue of variance due to covariate imbalance by randomly allocating treatment conditional on covariates believed to influence the outcome of interest. If we stratify in such a way that the probability an instance receives a given treatment is independent of its covariates, for example, by grouping instances by each assignment to the covariates and then assigning treatment randomly with fixed probabilities, the causal graphical model in figure \ref{fig:random_experiment_network} still holds. We can then estimate the average causal effects directly from the differences in outcome across treatments. More complex stratification strategies can introduce a backdoor path from treatment to outcome via the covariates on which treatment is stratified, (figure \ref{fig:random_experiment_network_stratified}). This necessitates that one condition on these covariates in computing the average casual effect in the same way as for estimating causal effects under ignorability (\S\ref{sec:causal_effect_under_ignorability}). The key difference is that the propensity score is known, as it is designed by the experimenter, and there are guaranteed (rather than assumed) to be no latent confounding variables (that influence both treatment and outcome). See \citet{imbens2015causal} for a discussion of the trade-offs between stratified versus completely random experiments.

@@

%\todo{discuss the key ways in which the simple regret problem differs from cummulative regret}

>>>>

There are many people without whose support this thesis would not have been possible. I would like to thank my supervisors Dr Cheng Song Ong, Dr Mark Reid and Dr Tiberio Caetano, Prof. Robert Williamson for chairing my panel, my brother Tor for many interesting and insightful discussions on bandit problems, and my parents for helping me proofread and the many days given to care for my lovely daughters. I would also like to thank Victoria, Inger, Natalie and all the people at ANU thesis bootcamp for giving me the techniques I needed to get past the blank page and get words out. Finally I want to thank my wonderful husband for his unwavering support and willingness to take on the household chaos, and my two beautiful daughters Freya and Anouk for their patience while mummy worked on her "tesis".

@@

\section{Motivation}
Many of the most important questions in science, commerce and our personal lives are about the outcomes of doing something. Will asking people to pay upfront at the doctors reduce long term health expenditure? If we developed a drug to suppress particular genes, could we cure multiple sclerosis? Would delaying teenage pregnancies improve the outcome for their children.

@@

The notion of causality has been widely debated in science and philosophy \citep{hume1741treatise,mill1893system,pearson1911grammar,Pearl2000,Lewis2000,woodward2005making,heckman20051,cartwright2007hunting} but is still viewed as poorly defined. This has led to a reluctance among applied researchers in many fields to make concrete claims about causality in their work, leading them instead to report that variables are \textit{related}, \textit{correlated} or \textit{associated}. However, the magnitude, direction and even existence of an association depends on which variables are controlled for (or included in a regression). Avoiding formalising causation, which is the real question of interest, requires the reader to determine via \quotes{common sense} the implications of the reported associations.

There are two ways in which an association detected in a data set may be non-causal. The first is that the variables concerned may not be related at all, and the association has arisen by chance in that data sample. Given finite data on enough variables, there is a high probability of finding some that appear correlated even though they are completely unrelated. For example, based on data from the years 1999 to 2009, the age of Miss America is strongly correlated with the number of murders  (in the US) by steam, hot vapours and hot objects \citep{Vigen2015}. We would not expect this relationship to hold in a new sample of data. This form of spurious correlation also has serious repercussions. It lies at the heart of major problems with the process of scientific research: researchers are incentivised to detect effects and thus to explore many possible paths in the process of analysing data and studies that fail to find an effect are less likely to be published. Consequently, the likelihood that reported effects have arisen by chance is underestimated, leading to the conclusion that \quotes{most published scientific results are false} \citep{Ioannidis2005}. This issue is also highlighted by recent crises in replication \citep{OpenScienceCollaboration2015}. This issue can be ameliorated by obtaining more data and by separating learning models from evaluating their performance, for example by evaluating models on a strict hold-out set or on the extent to which their results can be replicated.

@@

In chapter \ref{chap:causal-bandit}, I introduce causal bandit problems that unify causal graphical models and multi-armed bandit problems. Bandit arms are related to interventions in a causal graphical model in a very natural way: each multi-armed bandit arm (or action) corresponds to a particular assignment of values to variables within the causal graphical model. I show how the causal bandit framework can be used to describe a number of existing problems that lie in the intersection between the observational and interventional approaches to causality and demonstrate when causal bandit problems reduce to different existing bandit settings depending what information is observable and when.

In \S\ref{sec:causal-bandit-post-action-feedback}, I focus on causal bandit problems for which the values of variables in the causal graph are observed after an action is selected. I demonstrate that this leads to a novel form of structure between the bandit arms that cannot be exploited by existing bandit algorithms. In \S\ref{sec:simple-regret}, I describe and develop an algorithm for a special case of the causal bandit problem, which I refer to as the \emph{parallel bandit problem}. I demonstrate via upper and lower bounds on the regret that the algorithm is close to optimal for these problems and that the introduction of the causal structure leads to substantially faster learning. In \S\ref{sec:simple-regret-general}, I develop and prove regret bounds for an algorithm that can be applied to arbitrary causal graphs, albeit with stronger assumptions on what must be known a priori, and introduce a measure that captures the underlying difficulty of causal bandit problems, which depends on the causal graph and can be viewed as an \quotes{effective number of arms}. I also show how this measure can be used to quantify the value of optimised interventional data over purely observational data. Section \ref{sec:causal-bandit-experiments} provides experiments demonstrating the performance of both the parallel and general causal bandit algorithms on causal bandit problems. Finally in \S\ref{sec:causal_bandit_discussion}, I discuss extensions and potential future work.

@@

Counterfactuals \citep{Rubin1974} were developed from the starting point of generalising from randomised trials to less controlled settings. They describe causal effects in terms of differences between counterfactual variables, what would happen if we took one action versus what would happen if we took another. Counterfactual assertions can be expressed very naturally in human languages and are  prevalent in  everyday conversations; \quotes{if I had worked harder I would have got better grades} and \quotes{she would have been much sicker if she hadn't taken antibiotics}. Structural equation models have been developed and applied primarily within economics and related disciplines. They can be seen as an attempt to capture key aspects of the people's behaviour with mathematics. Questions around designing policies or interventions play a central role in economics. Thus they have transformed simultaneous equations into a powerful framework and associated set of methods for estimating causal effects. The is also a rich strand of work on using the assumptions that can be encoded in structural equation models, also known as functional causal models to discover the structure and direction of causal relationships - see for example \citet{mooij2016distinguishing}. Causal Bayesian networks \citep{Pearl2000} are a more recent development and arise from the addition of a fundamental assumption about the meaning of a link to Bayesian networks. They inherit and leverage the way Bayesian networks encode conditional independencies between variables to localise the impact of an intervention in a system in a way that allows formalisation of the conditions under which causal effects can be inferred from observational data.

An understanding of causal Bayesian networks and their properties (in particular the do calculus, see section \ref{sec:do_calculus}) is sufficient to appreciate my main technical contributions in chapter \ref{chap:causal-bandit}, as well as the importance of formal causal reasoning in Bayesian inference that I highlight in section  \ref{sec:bayesian_causality}. However, the literature on causal inference techniques remains split between the different frameworks. Much of the recent work on estimating causal effects within machine learning, as well as widely used methodologies such as propensity scoring, are described using the counterfactual framework. Methods developed within economics, in particular instrumental variable based approaches, or those requiring parametric or functional assumptions, are often based around structural equation models. This makes it worthwhile for researchers interested in causality to develop an understanding of all these viewpoints.

@@

\vspace*{.3cm}
\begin{example}
\label{exm:adjusting}
Suppose we want to estimate the impact on high school graduation rates of compulsory preschool for all four year olds. We have a large cross-sectional data set on a group of twenty year olds that records if they attended preschool, if they graduated high school and their parents socio-economic status (SES). We will let $X\in \set{0,1}$ indicate if an individual attended preschool, $Y \in \set{0,1}$ indicate if they graduated high school and $Z \in \set{0,1}$ represent if they are from a low or high SES background respectively.\footnote{There has been substantial empirical work on the effectiveness of early childhood education including a landmark randomised trial, the Perry Preschool project, which ran from 1962-1967 \citep{weikart1970longitudinal}.}
\end{example}

@@

Returning to example \ref{exm:ranomized_experiment}, and phrasing our query in terms of interventions; what would the distribution of outcomes look like if everyone was treated $\P{Y|do(X=1)}$, relative to if no one was treated $\P{Y|do(X=0)}$? The treatment $X$ is a potential cause of $Y$, along with other unobserved variables, such as the age, gender and the disease subtype of the patient. Since $X$ is assigned via deliberate randomisation, it cannot be affected by any latent variables. The causal Bayesian network for this scenario is shown in figure \ref{fig:causal_network_example}. This network represents the (causal) factorisation  $\P{X,Y} = \P{X}\P{Y|X}$, so from equation (\ref{eq:truncatedproduct}), $\P{Y|do(X)} = \P{Y|X}$. In this example, the interventional distribution is equivalent to the observational one.

@@

Putting aside any estimates of individual causal effects, it is possible to learn something about the distributions under treatment or placebo. Let $\cf{Y}{1}$ be a random variable representing the potential outcome if treated. The distribution of $\cf{Y}{1}$ is the distribution of $Y$ if everyone was treated. Similarly $Y^{0}$ represents the potential outcome for the placebo. The difference between the probability of recovery, across the population, if everyone was treated and the probability of recovery if everyone received the placebo is $\P{\cf{Y}{1}}-\P{\cf{Y}{0}}$. We can estimate (from an experimental or observational study):
\begin{itemize}
\item $\P{Y=1|X=1}$, the probability that those who took the treatment will recover
\item $\P{Y=1|X=0}$, the probability that those who were \emph{not} treated will recover
\end{itemize}

@@

The assumptions $X \ci Y^{1}$ and $X \ci Y^{0}$  are referred to as ignorability assumptions \citep{Rosenbaum1983}. They state that the treatment each person receives is independent of whether they would recover if treated and if they would recover if not treated. This is justified in example \ref{exm:ranomized_experiment} due to the randomisation of treatment assignment. In general the treatment assignment will not be independent of the potential outcomes. In example \ref{exm:adjusting}, the children from wealthy families could be more likely to attend preschool but also more likely to do better in school regardless, i.e $X \nci \cf{Y}{0}$ and $X \nci \cf{Y}{1}$. A more general form of the ignorability assumption is to identify a set of variables $Z$ such that $X \ci Y^{1}|Z$ and $X \ci Y^{0}|Z$.

@@

Assuming that within each socio-economic status level, attendance at preschool is independent of the likelihood of graduating high-school had a person attended, then the average rate of high-school graduation given a universal preschool program can be computed from equation \ref{eqn:counterfactual1}. Note, that this agrees with the weighted adjustment formula in equation \ref{eqn:backdoor_example}.

Another assumption introduced within the Neyman-Rubin causal framework is the Stable Unit Treatment Value Assumption (SUTVA) \citep{Rubin1978}. This is the assumption that the potential outcome for one individual (or unit) does not depend on the treatment assigned to another individual. As an example of a SUTVA violation, suppose disadvantaged four year olds were randomly assigned to attend preschool. The subsequent school results of children in the control group, who did not attend, could be boosted by the improved behaviour of those who did and who now share the classroom with them. SUTVA violations would manifest as a form of model misspecification in causal Bayesian networks.

@@

The SEM for example \ref{exm:adjusting} is shown in figure \ref{fig:sem:preschool}. Intervening to send all children to preschool replaces the equation $X = f_x(Z,\epsilon_x)$ with $X=1$, leaving all the other functions and distributions in the model unchanged.

@@

\vspace{.3cm}
\begin{theorem}
\textbf{(Local Markov condition)} Given a Bayesian network $G$ with nodes $X_{1}...X_{N}$, each variable $X_{i}$ is independent of its non-descendents given its parents in $G$ for all distributions $P(X_{1}...X_{N})$ that are compatible with $G$.
\end{theorem}

@@

In some cases, the average causal effect for some subgroup of the population is of prime interest. A particularly common example of this is the average treatment effect of the treatment of the treated (ATT). This would be the key quantity of interest if we had to decide whether or not to continue providing a program or treatment for which we could not control the treatment assignment process.

@@

Assuming $x$ and $\vb{z}$ are discrete, equation \ref{eqn:expected_post_intervention}, and thus the ACE, can be estimated by selecting the data for which $X=x$, stratifying by $\vb{Z}$, then computing the mean outcome within each stratum and finally weighting the results by the number of samples in each strata. However this approach is not workable for most real world problems with finite samples as the number of strata grows exponentially with the dimension of $\vb{Z}$ and it cannot handle continuous covariates. There is a substantial body of work within in the statistics and econometrics literature on estimating average causal effects assuming conditional ignorability (see \citet{Imbens2004} for a comprehensive review). The key approaches are based on matching on covariates, propensity score methods and regression. We now examine these approaches from a machine learning perspective.

@@

\subsubsection{Propensity scores}
If the parametric model is missspecified then the parameter that minimises the loss depends on the distribution from which the covariates $z$ are sampled. The model learned by ERM could perform very well in a validation set (which estimates the generalisation error over the observational distribution of $(x,z)$) but yield very poor estimates of the causal effect, see figure \ref{fig:covariate_shift_adjust}.

\begin{figure}[ht]
\centering
\includegraphics[scale=.48]{covariate_shift_estimation.pdf}
\caption{
Parametric regression may yield poor estimates of causal effects if the model is missspecified, even if the model fits well over the domain of the training data. In this example, $\P{Z|X=0} \sim N(\mu_0,\sigma_0^2)$ and $\P{Z|X=1} \sim N(\mu_1,\sigma_1^2)$ with little overlap in the densities. If $X=0$ then $Y \sim N(f_1(x)=sin(x),\sigma_y^2)$ and if $X=1$ then $Y \sim N(f_0(x) = \frac{1}{x+1},\sigma_y^2)$. We estimate $f_1(z)$ from the sample in which $X=1$ (green points) and $f_0(z)$ from the sample for which $X=0$ (blue points). In both cases the linear model is a good fit to the data. However, the resulting estimate of the causal effect is very poor for the lower values of $z$. } 
\label{fig:covariate_shift_adjust}
\end{figure}

@@

The estimator in equation \ref{eqn:importance_weighted_erm} is an example of a doubly robust estimator \citep{Robins1994a,Kang2007}. Doubly robust methods are asymptotically unbiased so long as either the regression model $h$ or propensity score $e$ are correctly specified \citep{robins1995semiparametric}.

@@

This formulation of the regression estimator highlights the missing data aspect of causal effect estimation. For each instance, the regression models are used to estimate the counterfactual outcome had the instance received the alternate treatment. Matching estimates the counterfactual outcome for an instance from the outcome of \emph{similar} instances that received a different treatment. \citet{abadie2002simple} analyse an estimator where both target and control instances are matched and the matching is done with replacement, let $j \in J_k(i)$ be the indices of the $k$ instances closest to $i$ by some metric $d(z_i,z_j)$ such that $x_i \neq x_j$.

@@

The significance of this should not be underestimated. Cross-validation has allowed applied machine learning to succeed with a very a theoretical approach on the basis that we can identify when a model is successful. With causal effect estimation there is no guarantee that a model that performs well at prediction (even out of sample) will accurately estimate the outcome of an intervention. \citet{Sugiyama2007} propose inverse propensity weighted cross validation for the covariate shift problem. There is relatively little theory on model selection for estimating the propensity score. To achieve asymptotically unbiased estimates, the covariates should satisfy the backdoor criterion. It is also known that conditioning on instrumental variables, which directly influence $X$ but not $Y$, increases variance without any reduction in bias and can increase bias if there are unmeasured confounding variables \citep{Wooldridge2009,Bhattacharya2012,Pearl2012b,Myers2011}. With doubly robust estimators, one could apply an iterative approach, fitting a propensity score model, using the results for inverse propensity weighted cross-validation of the regression model and then selecting covariates for the propensity model on the assumption the estimated regression function was correct.

@@

Suppose data is generated by the linear Gaussian Bayesian network below. The goal is to estimate the causal effect of $X$ on $Y$, that is identify the value of the coefficient $w_{yx}$. The variable $U$ is latent.

@@

where,

\eqn{
\beta &= w_{yz}+\frac{w_{yu}w_{zu}}{w_{zu}^2+\frac{v_z}{v_u}}, \text{ and,}\\
\epsilon &= v_y + w_{yu}^2v_u - \frac{v_u^2 w_{zu}^2 w_{yu}^2 }{v_z^2+v_u^2 w_{zu}^2}
}

@@

The previous sections focused on aspects of the problem of estimating the likely effect of an intervention from data gathered prior to making the intervention. There is an obvious alternative. Instead of trying to infer the outcome of an intervention from passive observations, one can intervene and see what happens. There are three key differences between observing a system and explicitly intervening in it. First, we determine the nature of the intervention and thereby control the data points used to estimate causal effects. Selecting data points optimally for learning is the focus of the optimal experimental design literature within statistics \citep{pukelsheim2006optimal} and the active learning literature in machine learning \citep{settles2010active}. Secondly, explicitly choosing interventions yields a perfect model of the probability with which each action is selected, given any context, allowing control over confounding bias. Finally, when we are intervening in a system we typically care about the impact of our actions on the system in addition to optimising learning. For example, in a drug trial, assigning people a sub-optimal treatment has real world costs. This leads to a trade-off between exploiting the best known action so far and exploring alternative actions about which we are less certain. This exploration-exploitation trade-off lies at the heart of the field of reinforcement learning \citep{sutton1998reinforcement}.

@@

We focus on a simpler class of problems within reinforcement learning known as multi-armed bandit problems. Multi-armed bandit problems also describe an agent aiming to maximise some reward by interacting with an environment. However, the reward is observed immediately after the action is selected and the environment is stateless. Future rewards generated by the  environment in response to the agent's actions do not depend on the previous actions of the agent. These assumptions are a reasonable approximation in settings where we face the same decision repeatedly with respect to many independent units or individuals and we have a good proxy for the outcome. In this chapter we describe the interventionalist approaches to learning to act for these types of problems, beginning with classical randomised experiments and then the, more adaptive, multi-armed bandit algorithms. We show how both observational causal inference and bandit algorithms can be viewed as generalisations of randomised experiments and highlight connections between the observational and interventional approaches to causal inference.

@@

Stratified randomised experiments address the issue of variance due to covariate imbalance by randomly allocating treatment conditional on covariates believed to influence the outcome of interest. If we stratify in such a way that the probability an instance receives a given treatment is independent of its covariates, for example, by grouping instances by each assignment to the covariates and then assigning treatment randomly with fixed probabilities, the causal graphical model in figure \ref{fig:random_experiment_network} still holds. We can then estimate the average causal effects directly from the differences in outcome across treatments. More complex stratification strategies can introduce a backdoor path from treatment to outcome via the covariates on which treatment is stratified, (figure \ref{fig:random_experiment_network_stratified}). This necessitates that one condition on these covariates in computing the average causal effect in the same way as for estimating causal effects under ignorability (\S\ref{sec:causal_effect_under_ignorability}). The key difference is that the propensity score is known, as it is designed by the experimenter, and there are guaranteed (rather than assumed) to be no latent confounding variables (that influence both treatment and outcome). See \citet{imbens2015causal} for a discussion of the trade-offs between stratified versus completely random experiments.

@@

Pure-exploration does not mean simply playing the arm with the widest uncertainty bounds. The goal is to be sure the arm we believe is optimal is in fact optimal at the end of the exploration period. This means we should focus exploration on arms which are plausibly optimal, creating a form of exploration-exploitation trade-off, albeit subtlety different to that for the cumulative regret.
